RL-PoWER, Evolving Policy Parameterization with Reinforcement Learning and its Application to Quadrupedal Walking Optimization

### 2012-06-10 paper_aimsa/abstract.tex
Edits toward a more polished paper

A reinforcement learning approach which can evolve the policy parameterization dynamically during
the learning process for quadruped robot gait . By gradually increasing the representational
power of the policy parameterization, it manages to find better
policies faster than HyperNEAT, a state-of-the-art evolutionary ANN algorithm that achieved the best performance in previous studies.

Expectation-Maximization-based reinforcement learning algorithm
is used. The test was done on a physical quadrupedal robot(QuadraTot). The results show that the evolving policy parameterization combined with expectation-maximization (RL PoWER)
outperforms substantially the HyperNEAT, by having
faster convergence and higher final
reward. According to the statistics collected, the RL PoWER achieves 16.3\%
advantage over the HyperNEAT, the best method tested on QuadraTot in previous studies.

>>>>

A reinforcement learning approach which can evolve the policy parameterization dynamically during
the learning process for quadruped robot gait \edit{???}. By gradually increasing the representational
power of the policy parameterization, it manages to find better
policies faster than HyperNEAT, a state-of-the-art evolutionary ANN algorithm that achieved the best performance in previous studies.

Expectation-Maximization-based reinforcement learning algorithm
is used. The test was done on a physical quadrupedal robot (QuadraTot). The results show that the evolving policy parameterization combined with expectation-maximization (RL PoWER)
outperforms substantially the HyperNEAT, by having
faster convergence and higher final
reward. According to the statistics collected, the RL PoWER achieves 16.3\%
advantage over the HyperNEAT, the best method tested on QuadraTot in previous studies.

### 2012-06-10 paper_aimsa/experimentalSetup.tex
Edits toward a more polished paper

The quadruped robot used was assembled from parts purchased online and
parts printed by the Objet Connex 500 3-D Printing System. The robot
actuation system consists of 5 AX-18+ Dynamixel servos and 4 AX-12+
Dynamixel servos: one inner joint with one AX-18+ servo and one outer
joint with AX-12+servo in each of the four legs, and one AX-18+ servo
at the center. To avoid the formerly reported AX-18+ servos are used
in this robot because of their stronger actuation power than that of
AX-12+. Each servo could be set to a position in the range [0, 1023]
by using PyDynamixel library, corresponding to a physical range [-120,
  +120]. Also, to prevent collisions with the robot body, the control
module filter out the commands to a safe range. This range was [-85,
  +60 ] for the inner leg servos, [-113 , +39 ] for the outer leg
servos, and [-28 , +28] for the center hip servo. In the studies of
this paper, tethered cables powered both the computer and the
servos. It measures approximately 39.5 centimeters from leg to
opposite leg in the crouch position showed in [pic].Our performance
metric was the displacement over the evaluation period of 12 seconds
for each. Same as \cite{yosinski2011evolving-robot-gaits}, the displacement was
measured using a Wii remote that was placed on the ceiling. Different
from the original model described in \cite{yosinski2011evolving-robot-gaits}, the quadruped robot was
equipped with a three-infrared-LED cluster on top rather than just
one. The reason for this setup is that when fierce gaits were
executed, the Wii remote loses tracking of the robot position due to
the limited visible angle of a single LED.  These three LEDs were
place tightly together to act as one signal emitter. Each LED was
tilted outwards in order to maximize the visible range. [pic] A
separate tracking server ran on the robot PC interacted with the Wii
remote via bluetooth by using the CWiid library.  A corresponding
client communicates with this server via a socket connection. Our
fitness evaluation code talks with the Wii remote by using this client
and updates the position from beginning to end during each run. The
metric for evaluating gaits was the Euclidian distance the robot
traveled during a 12-second run on flat surface. Previous work done in
\cite{yosinski2011evolving-robot-gaits} and \cite{clune2009evolving-coordinated-quadruped} suggested that extremely fierce gaits are not viable in
general. These gaits tend to either overburden the servos or flip the
robot. Thus, RL\_POWER(?) were tested after carefully cropping out the
extremely fierce gaits each time. (As described earlier, each leg has
two joints, inner joint j1 and outer j2. The position of one leg is
determined by the sum of the values of j1 and j2. Extremely fierce
gaits usually have relatively small values of g. Thus cropping works
by mapping the wild gaits onto the boundary of a triangular space over
the two dimensional space of j1 and j2. \figref{smartCrop}....

\figp{smartCrop}{1}{$g = j1 +  j2$ caption}

\figp{sixLegPositions}{1}{caption... the left two are allowed, middle
  two are on the red border line from \figref{smartCrop}, right two
  are disallowed.}

>>>>

The quadruped robot used was assembled from parts purchased online and
parts printed by the Objet Connex 500 3-D Printing System. The robot
actuation system consists of 5 AX-18+ Dynamixel servos and 4 AX-12+
Dynamixel servos: one inner joint with one AX-18+ servo and one outer
joint with AX-12+servo in each of the four legs, and one AX-18+ servo
at the center. To avoid the formerly reported \edit{???} AX-18+ servos are used
in this robot because of their stronger actuation power than that of
AX-12+. Each servo could be set to a position in the range [0, 1023]
by using pydynamixel library, corresponding to a physical range [-120$^\circ$,
  +120$^\circ$]. Also, to prevent collisions with the robot body, the control
module filter out the commands to a safe range. This range was [-85$^\circ$,
  +60$^\circ$] for the inner leg servos, [-113$^\circ$, +39$^\circ$] for the outer leg
servos, and [-28$^\circ$, +28$^\circ$] for the center hip servo. In the studies of
this paper, tethered cables powered both the computer and the
servos. It measures approximately 39.5 centimeters from leg to
opposite leg in the crouch position showed in \figref{quadratotWhiteBooties}.

Our performance metric was the displacement over the evaluation period of 12 seconds
for each. Same as \cite{yosinski2011evolving-robot-gaits}, the displacement was
measured using a Wii remote that was placed on the ceiling. Different
from the original model described in \cite{yosinski2011evolving-robot-gaits}, the quadruped robot was
equipped with a three-infrared-LED cluster on top rather than just
one. The reason for this setup is that when fierce gaits were
executed, the Wii remote loses tracking of the robot position due to
the limited visible angle of a single LED.  These three LEDs were
place tightly together to act as one signal emitter. Each LED was
tilted outwards in order to maximize the visible range. A
separate tracking server ran on the robot PC interacted with the Wii
remote via bluetooth by using the CWiid library.  A corresponding
client communicates with this server via socket connection. Our
fitness evaluation code talks with the Wii remote by using this client
and updates the position from beginning to end during each run. The
metric for evaluating gaits was the Euclidian distance the robot
traveled during a 12-second run on flat surface. Previous work done in
\cite{yosinski2011evolving-robot-gaits} and \cite{clune2009evolving-coordinated-quadruped} suggested that extremely fierce gaits are not viable in
general. These gaits tend to either overburden the servos or flip the
robot. Thus, RL PoWER were tested after carefully cropping out the
extremely fierce gaits each time. As described earlier, each leg has
two joints, inner joint j1 and outer j2. The position of one leg is
determined by the sum of the values of j1 and j2. Extremely fierce
gaits usually have relatively small values of $g$. As shown in \figref{smartCrop}, this cropping method works
by mapping the wild gaits onto the boundary of a quasi-triangular area in
the two-dimensional space of j1 and j2.

\figp{smartCrop}{1}{The red line is the cropping border, $g = j1 +  j2$. As shown in \figref{sixLegPositions}}

\figp{sixLegPositions}{1}{The left two positions are allowed; middle
  two positions are on the red border line from \figref{smartCrop}; right two positions are deemed as ``extremely fierce'', thus are disallowed.}

### 2012-06-10 paper_aimsa/introduction.tex
Edits toward a more polished paper

Various learning algorithms have been proved to be effective for
legged robots. Algorithms such as HyperNEAT \cite{yosinski2011evolving-robot-gaits},
GA \cite{chernova2004an-evolutionary-approach-to-gait} and others \cite{hornby2005autonomous-evolution-of-dynamic} \cite{zykov2004evolving-dynamic-gaits} \cite{tellez2006evolving-the-walking-behaviour} \cite{valsalam2008modular-neuroevolution-for-multilegged} have been tested to be effective for automatic learning gaits for robots. Despite of the competetive performance, a major
task that are usually hidden in the published results lies in tuning the parameters for
these evolutionary algorithms \cite{kormushev2011bipedal-walking-energya}. Here we present a
different way for learning gaits using reinforcement learning and expectation maximization, called RL PoWER. In our
experiment, the main focus of the research is on the applicability of RL PoWER to quadruped robot gait learning. Another motivation is to compare the state-of-the-art neural network algorithm, HyperNEAT, with our proposed method in quadruped robot gait learning.

\subsection{Problem Definition}
The gait learning problem is defined to find a
gait that maximizes some specific metric. Mathematically, we define a
gait as a function that specifies a vector of commanded motor
positions for a robot over time. Gaits without feedback---also
called open-loop gaits---can be defined as

@@

According to this definition, open-loop gaits are deterministic. One
particular gait should behave exactly the same when it is run from
trial to trial. However, the actual robot motion and fitness measured
will vary due to the errors and uncertainty of the real world physics. In our
trials, the gaits generated were sent to the robot and
executed in an open loop manner. We will measure and analyze the performance between
RL PoWER and HyperNEAT, the latter of which will be the focus of discussion in this paper. The metric used for the fitness in this paper will be
described later.

>>>>

Various learning algorithms have been proved to be effective for
legged robots. Algorithms such as HyperNEAT \cite{yosinski2011evolving-robot-gaits},
GA \cite{chernova2004an-evolutionary-approach-to-gait} and others \cite{hornby2005autonomous-evolution-of-dynamic} \cite{zykov2004evolving-dynamic-gaits} \cite{tellez2006evolving-the-walking-behaviour} \cite{valsalam2008modular-neuroevolution-for-multilegged} have been tested to be effective for automatic learning gaits for robots. Despite competetive performance, a major
task that are usually hidden in the published results lies in tuning the parameters for
these evolutionary algorithms \cite{kormushev2011bipedal-walking-energy}. Here we present a
different way for learning gaits using reinforcement learning and expectation maximization, called RL PoWER. In our
experiment, the main focus of the research is on the applicability of RL PoWER to quadruped robot gait learning. Another motivation is to compare the state-of-the-art neural network algorithm, HyperNEAT, with our proposed method in quadruped robot gait learning.

\subsection{Problem Definition}

As in \cite{yosinski2011evolving-robot-gaits}, the gait learning problem is defined to find a
gait that maximizes some specific metric. Mathematically, we define a
gait as a function that specifies a vector of commanded motor
positions for a robot over time. Gaits without feedback---also
called open-loop gaits---can be defined as

@@

According to this definition, open-loop gaits are deterministic. One
particular gait should behave exactly the same when it is run from
trial to trial. However, the actual robot motion and fitness measured
will vary due to the errors and uncertainty of the real world physics. In our
trials, the gaits generated were sent to the robot and
executed in an open loop manner. We will measure and analyze the performance between HyperNEAT and 
RL PoWER, the latter of which will be the focus of discussion in this paper. The metric used for the fitness in this paper will be
described later.

### 2012-06-10 paper_aimsa/methods.tex
Edits toward a more polished paper

\subsection{Policy Representation by Splines}
The simplest model with back-compatibility is geometric
splines. For a given model f(x) with K knots, we can preserve the
exact shape of the generated curve while adding extra knots to the
original spline. Say, if we put one additional knot between every two
consecutive knots of the original spline, we end up with a 2K - 1
knots and a spline that has the same shape as the original one. In
order to do this, we need to define an algorithm for evolving the
parameterization from $K$ to $L$ knots ($L > K$), which is formulated as 
Algorithm 1 in \cite{kormushev2011bipedal-walking-energy}.  Without loss of generality, the policy parameters are
normalized into $[0, 1]$, and appropriately scaled/shifted as necessary
later upon use.

@@

For the experiment, we set 3 knots for each servo and there are 8
servos in total. The servo in the hip is not used in
our experiment. Previous work has verified that quadruped gaits
perform better when they are coordinated \cite{clune2009evolving-coordinated-quadruped} \cite{clune2011on-the-performance-of-indirect-encoding}
\cite{valsalam2008modular-neuroevolution-for-multilegged}. For each spline, we calculate its corresponding parameterized gait for one unit time cycle. Given that, then apply the same
pattern to every cycle throughout the 12 seconds of one
trial. Specifically, each spline(a set of 3 knots) is interpreted to its corresponding servo positions as
following:

\begin{table}[b]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameters        &                           &       \\
in $\vec{\theta}$ & Description               & Range \\
\hline
\hline
$f(s1,s2,s3)$        & Spline function           & [0,1] \\  %subject to change
\hline
$R$          & Position multiplier                & [256, 768] \\
\hline
\end{tabular}
\caption{The \emph{RL PoWER} motion model parameters.}
\tablabel{parameters}
\label{tab:params}
\end{center}
\end{table}

\[
\vec{g}(t) =
\left[ {\begin{array}{c@{ }c@{ }c@{ }l@{ }l}
R \cdot f(s1, s2, s3) & \ \          & \             & \            & + C \\ % 1
0                              & \             & \             & \            & + C_C \\ % 8
\end{array} } \right]
\]

>>>>

\subsection{Policy Representation by Splines}
The simplest model with back-compatibility is geometric
splines. \editbox{Too fast... add a sentence or two on what the methods are trying to do, and what back-compatibility is.} For a given model f(x) with K knots, we can preserve the
exact shape of the generated curve while adding extra knots to the
original spline. Say, if we put one additional knot between every two
consecutive knots of the original spline, we end up with a 2K - 1
knots and a spline that has the same shape as the original one. In
order to do this, we need to define an algorithm for evolving the
parameterization from $K$ to $L$ knots ($L > K$), which is formulated as 
Algorithm 1 in \cite{kormushev2011bipedal-walking-energy} \edit{reproduce that algorithm here?}.  Without loss of generality, the policy parameters are
normalized into $[0, 1]$, and appropriately scaled/shifted as necessary
later upon use.

@@

For the experiment, we set 3 knots for each servo and there are 8
servos in total. The servo in the hip is not used in
our experiment. Previous work has verified that quadruped gaits
perform better when they are coordinated \cite{clune2009evolving-coordinated-quadruped, clune2011on-the-performance-of-indirect-encoding, valsalam2008modular-neuroevolution-for-multilegged}. For each spline, we calculate its corresponding parameterized gait for one unit time cycle. Given that, then apply the same
pattern to every cycle throughout the 12 seconds of one
trial. Specifically, each spline (a set of 3 knots) is interpreted to its corresponding servo positions as shown in \eqnref{rl} and \tabref{params}.

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameters        &                           &       \\
in $\vec{\theta}$ & Description               & Range \\
\hline
\hline
$f(s1,s2,s3)$        & Spline function           & [0,1] \\  %subject to change
\hline
$R$          & Position multiplier                & [256, 768] \\
\hline
\end{tabular}
\caption{The \emph{RL PoWER} motion model parameters.}
\tablabel{parameters}
\tablabel{params}
\end{center}
\end{table}

\be
\vec{g}(t) =
\left[ {\begin{array}{c@{ }c@{ }c@{ }l@{ }l}
R \cdot f(s1, s2, s3) & \ \          & \             & \            & + C \\ % 1
0                              & \             & \             & \            & + C_C \\ % 8
\end{array} } \right]
\eqnlabel{rl}
\ee

\editbox{Add a new equation showing all servos (all nine) with their
  actual values of the parameters $R$ and $C$ plugged in.}

### 2012-06-10 paper_aimsa/relatedWork.tex
Edits toward a more polished paper

Other attempts to solve it

>>>>

Many attempts have been made on robot gaits learning using machine
learning algorithms, producing competitive results \edit{Change these
  all to citations} (Chernova and Veloso, 2005; Hornby et al., 2005;
Zykov et al., 2004; Clune et al., 2009a, 2011, 2009b,c; T´ellez et
al., 2006; Valsalam and Miikkulainen, 2008). In fact, Jason et
al. tested and compared six different algorithms for quadruped robot
gait learning in their studies. HyperNEAT, a generative encoding
neural network algorithm, achieved the best performance among all six
methods. In Kormushev et al. (2011)'s bipedal robot energy reduction
research, a reinforcement learning method was used to optimize the
performance. The results from Kormushev's research showed this
algorithm's potential for walking problems. This reinforcement
learning method still needs more tests for further explorations.

### 2012-06-10 paper_aimsa/results.tex
Edits toward a more polished paper

\section{Results}

RL\_POWER Gaits

The results for the gaits evolved by RL PoWER are shown in Figure 8
and Table 2. A total of 900 evaluations were performed for HyperNEAT
(300 in each of three runs). Overall the RL PoWER gaits were the
faster and more stable by far, beating the HyperNEAT when comparing
either average or best gaits.

(pic)RL POWER AND HyperNEAT performance chart

We believe that this is because RL PoWER uses evolvable splines to represent gaits As explained earlier,  
The single best gait found during this study had a speed of

11.0855923766 centimeters/sec, XX\%

better than the best HyperNEAT gait. Figure 9 shows a typical
RL\_POWER gait that had high fitness. The pattern of motion is less
complex and more regular than the HyperNEAT, in that patterns of
multiple motors are coordinated.

A corresponding observation of this property is that the noisiness of
RL\_POWER is significantly lower than that of HyperNEAT. One reason of
this is that the when Expectation Maximization methodology is used,
explorations happened in the vicinity space of the highest ranked
patterns. Even some individual HyperNEAT gaits turned out very
effective, the average noisiness of HyperNEAT is very large. For
example, as found in [xx], the mean of HyperNEAT generation-champion’s
performance was XX body lengths/minute (+-XX SD), but it had a max of
38 and a min of 3. RL\_POWER, on the other hand, proved to be stable
and much less noisy throughout the entire experiment. The maximum SD
over all three runs is +- XX SD. This may also be due to a much larger
space of possibilities of HyperNEAT than the more heuristically guided
reinforcement learning.

[pic][speed chart of two algos]

>>>>

\section{Results}
The results for the gaits evolved by RL PoWER are shown in \figref{rlPerform} and \tabref{results}.The HyperNEAT performance is shown in \figref{hyperPerform}. A total of 900 evaluations were performed for RL PoWER (300 for each of three runs). The reason these two algorithms have different number of trials is that runs continued until the performance plateaued, which we defined as when there was no improvement during the last third of a run. Overall, the RL PoWER gaits were
faster by far, beating the HyperNEAT when comparing
either average or best gaits.

\figp{rlPerform}{0.95}{Average fitness ($\pm$ SE) in cm/sec of the highest performing run of RL PoWER's three runs. }
\figp{hyperPerform}{0.75}{Average fitness ($\pm$ SE) in cm/sec of the highest performing
individual in one HyperNEAT run.}

\begin{table}
\begin{center}
\begin{tabular}{|r|c|c|c||c|}
\hline
                                         & Average & Std. Dev \\
\hline                                    
\hline                                    
RL PoWER                       & 7.62  &    2.1   \\
\hline
HyperNEAT                      & 6.28   &   1.26   \\
\hline
\end{tabular}
\caption{The average and standard deviation of the best gaits found
  for both algorithm during each of three runs, in cm/sec.}  \tablabel{results}
\end{center}
\end{table}

We believe that this is because RL PoWER uses evolvable splines to represent gaits As explained earlier,  
The single best gait found during this study had a speed of 11.09 centimeters/sec,  14.7\% better than the best HyperNEAT gait. \figref{rlPattern} shows a typical RL PoWER gait that had high fitness. Compared with \figref{neat_110115_211410_00000_002_filt_zoom}, the pattern of RL PoWER's motion is less
complex but more regular than the HyperNEAT, but producing higher speed. As shown in \figref{rlPattern}
, multiple motors are more coordinated than the HyperNEAT gait. 
\figp{rlPattern}{0.75}{The motor positions of a typical well performing gait over one cycle. It matches exactly the shape of its corresponding spline representation. This pattern shows the simplicity and power of geometric splines representations.}

\figp{neat_110115_211410_00000_002_filt_zoom}{0.75}{A typical well performing HyperNEAT gait's pattern of motor positions}

A corresponding observation from the graph on this property is that the noisiness of
RL PoWER is higher than HyperNEAT. One reason for this is that the RL PoWER intentionally adds noise when exploring the space near the best ranked policies. Another phenomenon seen from the \figref{rlPerform} is that one run of RL PoWER's general performance largely depends on initial gaits, which is a trait of hill climbing algorithms. The larger standard deviation in RL PoWER for space exploring can be easily fixed by tuning the noise parameter. RL PoWER converges to optimality at a higher rate. This is due to the more heuristically guided reinforcement learning of RL PoWER. HyperNEAT, on the other hand, has a larger dimension to explore due to its generative encoding method. While the evolvable spline representation used by RL PoWER has lower dimensions. The convergence of HyperNEAT is thus slower.
